---
title: "Introduction to survival analysis"
output:
  ioslides_presentation:
    widescreen: true
    css: ../styles/styles.css
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

# Overview of key ideas in survival analysis

## Studying long-term disease

We enroll people in a study with and without an exposure (history of obesity)

We want to see whether this exposure affects development of an outcome of interest (cancer)

What complications may arise when collecting data?

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = 0, y0 = 5:3, x1 = 1, lwd = 4)
points(x = c(0,0,0,1), y = 5:2, pch = 16, cex = 2)
abline(v = c(1,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1), y0 = 5:2, x1 = 2, lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
abline(v = c(2,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = 3, lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
abline(v = c(3,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,rep(4,4)), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = rep(4,2), y = c(4,3), pch = "X", cex = 2)
abline(v = c(4,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,5,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(5,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,6,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(6,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,7,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(7,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,8,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = c(3,8), y = c(5,2), pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(8,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,8,5), lwd = 4)
segments(x0 = 8, y0 = 2, x1 = 10, lwd = 4, lty = "dashed", col = "gray50")
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = c(3,8), y = c(5,2), pch = 1, cex = 2)
points(x = c(rep(4,2),5,10), y = c(4,3,1,2), pch = "X", cex = 2)
abline(v = c(8,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

## Time-to-event data

Studies in which patient outcomes are followed over time tend to use analysis methods in a statistical area known as **survival analysis**.

Key feature: outcome is the time to an event. This time is called a **survival time**, and the event is typically called a **failure**.

- One event is of primary interest -> traditional survival analysis
- Multiple events of interest -> competing risks or recurrent event analysis
- Data collected from such studies almost always exhibit **censoring**.
- Important idea we will return to: how should time be defined? Age, time since study began?

## Examples of time-to-event data

- Time to hospital readmission
- Time to reincarceration
- Time to understanding of a concept

## Survival analysis vs. other approaches

Why not treat failure as a binary event and use logistic regression?

- Ignores time information

Wny not treat the time-to-event as a continous variable and use linear regression?

- Ignores censoring information

## Goals of survival analysis

- Estimate and compare survival over time (survival functions) for groups of interest
- Estimate the effect of covariates on survival

# Censoring

## Types of censoring

- Right-censoring
- Left-censoring
- Interval-censoring
- Type I censoring
- Type II censoring

## Right-censoring

True survival time is greater than or equal to the observed survival time.

- Data collection stops at a prespecified time.
- Subject withdraws from study or is lost to follow-up before the failure.

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,8,5), lwd = 4)
segments(x0 = 8, y0 = 2, x1 = 10, lwd = 4, lty = "dashed", col = "gray50")
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = c(3,8), y = c(5,2), pch = 1, cex = 2)
points(x = c(rep(4,2),5,10), y = c(4,3,1,2), pch = "X", cex = 2)
abline(v = c(8,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
text(x = c(1.75,6.75), y = c(4.75,1.75), labels = "Right-censored", col = "red")
```

## Left-censoring

True survival time is less than or equal to the observed survival time.

- Disease detection occurs after physiological onset. e.g. Flu infection has occurred before you start to get a fever.

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,2), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = 0, x1 = 7, y0 = 1, lwd = 4)
points(x = c(5,7), y = c(1,1), pch = "X", cex = 2, col = c("red", "black"))
text(x = c(5,7), y = c(0.75,1.25), labels = c("Infection onset", "Infection detection"), col = c("red", "black"))
```

## Interval censoring

True survival time is between two known times

- Disease is detected between regular follow up visits

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,2), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = 0, x1 = 7, y0 = 1, lwd = 4)
points(x = c(3,5,7), y = rep(1,3), pch = "X", cex = 2, col = c("black", "red", "black"))
text(x = c(3,5,7), y = c(1.25,0.75,1.25), labels = c("No infection detection", "Infection onset", "Infection detection"), col = c("black", "red", "black"))
```

## Type I censoring

True survival time is greater than a pre-determined cutoff

- Special case of right censoring
- Also called administrative censoring (administrative end of the study)

## Type II censoring

Study stops when the $k$ subjects experience failure. The remaining subjects are censored.

- Can occur in product testing

## Censoring in WHI study

What types of censoring mechanisms might be at work in the Women's Health Initiative study on obesity duration and cancer risk?

## Assumptions about censoring

Typical survival analyses make the following assumptions regarding censoring:

- Random
- Independent
- Non-informative

## Random censoring

- Survival times are independent of censoring times.
- "Subjects who are censored at time $t$ should be representative of all the study subjects who remained at risk at time $t$ with respect to their survival experience." (Kleinbaum & Klein 2012)
- Split your subjects randomly into any number of groups. Within each of these groups, we expect the failure rates to be the same in the censored and uncensored individuals.
    - We don't know the actual failure rate in the censored individuals, but we **assume** it is the same as for the uncensored individuals.
- Stronger assumption than independent censoring

## Independent censoring

- If we have random censoring within each level of any **covariate**, we say that we have independent censoring.
- "Within any subgroup of interest, the subjects who are censored at time $t$ should be representative of all the subjects in that subgroup who remained at risk at time $t$ with respect to their survival experience." (Kleinbaum & Klein 2012)
- Random censoring implies independent censoring but not vice-versa.
- e.g. Covariate: sex.
    - Within males, we assume censored and uncensored individuals have the same failure rate. Same for within females.
    - Having the previous two facts does not imply that censored and uncensored individuals have the same failure rates in any random group.

## Non-informative censoring

- Occurs if the survival times don't provide information about the censoring times and vice-versa
- "The assumption of non-informative censoring is often justifiable when censoring is independent and/or random; nevertheless, these three assumptions are not equivalent." (Kleinbaum & Klein 2012)
- Usually we worry about informative censoring because the characteristic that drives the informativeness typically also results in the breakdown of independent censoring.
    - e.g. Existing medical condtions as informative variable
        - Existing heart disease affects both censoring times and failure times (informative)
        - Patients with existing heart disease tend to be on placebo and have worse survival outcomes than those without existing heart disease (non-independent)

## Example: random and independent, but informative

When a subject experiences the failure event, another subject drops out. $\Rightarrow$ informative censoring

Assume that this dropout subject's survival probability is representative of those who stayed in the study. $\Rightarrow$ independent and random censoring

## Importance of independent censoring

Usually survival analyses compare the survival of two or more groups. In this case, independent censoring allows for unbiased estimates of these differences. Why?

ex) Clinical trial of a new drug vs placebo

- Placebo users might drop out if they do not see their health changing.
- These censored subjects may be less healthy than the other subjects.
- If we assume independent censoring, we would assume that these censored subjects are just as healthy as the uncensored subjects (the ones on the new drug).
- This could cause us to underestimate the effect of the drug.

## Censoring assumptions in the WHI study

Are there any subject characteristics that might break the independent censoring assumption? In what direction would the effects be biased if this were the case?

# Statistical analysis

## Overview

- Estimating the survival function
    - Kaplan-Meier estimation
- Comparing Kaplan-Meier curves
    - Log-rank tests and their variations
- Estimating the effect of covariates on survival outcomes
    - Cox proportional hazards regression
    - Parametric models of survival

## Notation and terminology

$T_i$: observed survival time for subject $i$ (random variable)

$\Delta_i$: censoring indicator for subject $i$ (1 = censored, 0 = observed)

The probability distribution function for $T$ is $f(t)$, and the cumulative distribution function is $F(t) = P(T \geq t)$.

Data in a survival analysis setting include $T_i, \Delta_i$ and covariates.

## Functions of interest

Survivor function: probability of survival past a time $t$

$$ S(t) = P(T > t) = 1 - F(t) $$

Hazard function: instantaneous rate of failure at time $t$ given survival up to time $t$

$$ h(t) = \lim_{\Delta t \to 0} \frac{P(t \leq T < t+\Delta t \mid T \geq t)}{\Delta t} $$

## Visualizing functions of interest

```{r fig.width=10}
par(mfrow = c(1,2))
t <- seq(0, 20, 0.01)
s1 <- pexp(t, lower.tail = FALSE, rate = 0.1)
s2 <- pexp(t, lower.tail = FALSE, rate = 0.2)
plot(t, s1, type = "l", ylim = range(s1, s2), xlab = "Time", ylab = "Survival probability")
lines(t, s2, col = "red")
text(x = c(5,15), y = c(0.05,0.45), labels = c("Worse survival", "Better survival"), col = c("red", "black"))
## Hazard
plot(t, rep(0.1, length(t)), type = "l", xlab = "Time", ylab = "Hazard", ylim = c(0,0.3))
lines(t, rep(0.2, length(t)), col = "red")
text(x = c(10,10), y = c(0.07,0.23), labels = c("Better survival", "Worse survival"), col = c("black", "red"))
```

## Visualizing hazard functions

```{r fig.width=10}
par(mfrow = c(2,2))
h1 <- rep(3, length(t))
h2 <- exp(t-10)
h3 <- exp(-t-2)
h4 <- sin(t*1.5*pi/20)
plot(t, h1, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Healthy individuals")
plot(t, h2, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Cancer patients")
plot(t, h3, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Newborn infants")
plot(t, h4, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Acute onset, long-term disease")
```

## Goal: estimate and compare functions among groups

```{r fig.width=10}
library(survival)
data(lung)
lung$SurvObj <- with(lung, Surv(time, status == 2))
km.by.sex <- survfit(SurvObj ~ sex, data = lung, conf.type = "log-log")
t <- seq(0,1000)
s_true1 <- pexp(t, lower.tail = FALSE, rate = 0.002)
s_true2 <- pexp(t, lower.tail = FALSE, rate = 0.004)
par(mfrow = c(1,2))
plot(t, s_true1, type = "l", xlab = "Time", ylab = "Survival probability", main = "True survival functions", ylim = c(0,1))
lines(t, s_true2, col = "red")
legend("topright", legend = c("group 1", "group 2"), col = c("black", "red"), lty = "solid", bty = "n")
plot(km.by.sex, mark.time = FALSE, col = c("red", "black"), xlab = "Time from study enrollment", main = "Estimated survival functions")
```

## Estimating survival function

- The **Kaplan-Meier method** is the most common method used to estimate survival functions.
    - Take every event time in the dataset (either a failure time or a censoring time) and estimate $S(t)$ at each of these time points by looking at "risk sets" at these time points
    - Intuition for the estimator: what is the probability of surviving past time $t$ given that you've survived up to time $t$?

$$ \hat S(t_{(f)}) = \prod_{i=1}^{f} \hat P(T > t_{(i)} \mid T \geq t_{(i)}) $$

## Comparing survival functions

- Often we are interested in comparing survival trajectories in different subgroups. e.g. between treatment and placebo, genders, age groups
- The **log-rank test** can be used to test for a difference between two or more survival curves
    - $H_0$: no difference between the survival curves
    - How does the test work?
        - Recall chi-squared and Fisher's exact test for testing if two factors are independent
        - We want to see if group indicator is independent of survival at each time point

## Example output: log rank test

```{r}
survdiff(SurvObj ~ sex, data = lung)
```

## Estimating effects of covariates on survival

- The **Cox proportional hazards model** is a regression technique suitable for modeling survival data
    - Familiar features with logistic regression
        - The (natural) log is applied to the outcome on the left hand side. In the Cox model, the left hand side is the log hazard rate at a time $t$ as opposed to the log odds in logistic regression.
        - Coefficients must be exponentiated in order to obtain natural scale interpretations. In the Cox model, exponentiated coefficients are **hazard ratios** while in logistic regression, they are odds ratios.
        - $h_0(t)$ is called the **baseline hazard** and does not need to be specified. Our interest is in differences (ratios) between hazards indicated by $\beta$'s

$$ \log h(t; x_1,\ldots,x_p) = \log h_0(t) + \sum_{i=1}^p \beta_i x_i $$

## Cox model with time-varying covariates

- If all covariates are **time-independent** (do not change over time), then the **proportional hazards** assumption can be met.
    - Hazard functions predicted by the model for any two people will be proportional.
    - In practice, need to check that hazards in different groups are proportional.
- If time-dependent covariates are used, then the proportional hazards assumption is not met, but the **extended Cox model** can be used instead.
- This model assumes that the hazard at time $t$ depends on the time dependent covariates exactly at time $t$ and not a later or earlier time.
    - This can be addressed by "lagging" the time dependent covariates. e.g. $x_1(t-1)$ as opposed to $x_1(t)$.

$$ \log h(t; \mathbf{X}) = \log h_0(t) + \sum_{i=1}^{p_1} \beta_i x_i + \sum_{j=1}^{p_2} \beta_j x_j(t) $$

## More details on the Cox model

- Typically in regression situations we have seen so far, we must assume that the outcomes follow a particular probabilty distribution - **parametric approaches**.
    - Linear regression: outcomes come from a normal distribution
    - Logistic regression: outcomes (binary) come from a Bernoulli/binomial distribution
    - Poisson regression: outcomes come (counts) come from a Poisson distribution
- No such assumptions need to be made for the Cox model in order to estimate coefficients and standard errors
- The Cox model can also be used to model recurrent events such as relapse, tumor formation

## Example output: Cox model

```{r}
coxph(SurvObj ~ age + sex, data = lung)
```

## Parametric approaches

- Parametric models assume that the survival times follow a particular probability distribution
    - Weibull, exponential models
        - Covariates relate to the mean of these distributions
    - Accelerated failure time models
        - Outcome is failure time and is modeled as a function of covariates

# Data quality and bias

## Immortal time bias

- With survival analysis, we must worry about factors influenced by time
- Immortal time bias occurs when there is a time span for which a set of individuals could not possibly have experienced the failure event and are thus not comparable with other individuals
- e.g. Comparing transplantation to no transplantation
    - Failure event of interest: death
    - 80 patients originally. 40 die before having access to organs. Of the 40 remaining, 20 gain access to organ donations and have the transplant.
    - If we group the initial 40 who die before access with the "no transplant" group, we could overestimate the effect of transplantation.
- Can avoid this bias by carefully considering/defining the time variable and only comparing comparable individuals

## Lead time bias

- Lead time bias occurs if a screening procedure tends to occur before the usual time of diagnosis
- e.g. Differential screening between treatment arms
    - Failure event of interest: death
    - Say that all individuals die 10 years after disease onset. No difference in survival between treatment arms.
    - If time from disease detection until death is used and if the treatment group tends to be screened more aggressively, the treatment group will have longer survival (longer time from detection to death)
- Avoid this by selecting a suitable time metric and standardizing screening procedures

## Length bias

- Conditions that have slower progression or longer duration are more likely to be detected with the follow-up structure of long term studies
- Can be a concern if comparing survival among different conditions
- Hard to avoid but be aware when designing the study and reporting results

## Other issues

- Assess if there was differential selection into groups being compared
- Were measurement practices uniform over time?
- Are there exposures that may have changed over time that could affect the results?

# Competing risks

## Overview

- Sometimes when following a cohort to observe an outcome of interest, other outcomes can "compete" to occur first. These other outcomes are called **competing risks**.
    - e.g. If illness is the endpoint of interest, death or an unrelated illness may compete to occur first
    - If a competing risk occurs before the endpoint of interest, should we treat the individual as censored?
- Goals for competing risks analysis
    - What is the effect of covariates on the survival probability associated with any one of the competing risks?
    - Compare survival probabilities for competing risk A and competing risk B adjusting for covariates

## Statistical approaches

- **Cause-specific Cox models**: perform a Cox analysis for each competing risk, treating the other competing risks as censored observations
- **Cumulative incidence curves** give the marginal probability of experiencing a particular competing risk over time (i.e. probability of experienceing competing risk A regardless of what other competing risks may have occurred)
    - Methods exist to model CICs with covariates assuming proportional hazards
- **Conditional probability curves** give the probability of experiencing a particular competing risk event by time $t$ given that an individual has not experienced any of the other competing risks by time $t$.
    - Methods exist to compare two or more CPC curves

## References

David G Kleinbaum, Mitchel Klein. Survival Analysis: A Self-Learning Text. 3rd Edition. Springer. 2012.