---
title: "Review worksheet"
output: pdf_document
---

1. The sensitivity and specificity of a diagnostic test are both 80%. Calculate the positive and negative predictive value of this test for a prevalence of (a) 10% and (b) 1%. What do you notice? Can you come up with an analogy for hypothesis testing? In other words, can you translate type I error rate and power into this diagnostic testing context? What is an analogy for prevalence in the hypothesis testing setting?

2. The following logistic regression model is used to obtain propensity scores:

$$ \hbox{log odds of treatment} = -0.5 + 0.01\mathrm{age} + 0.05\mathrm{male} $$

What is the propensity score for a 30 year old female and for a 25 year old male?

3. What is going on in Parts 1-4 in the output below?

Example 1

```{r echo=FALSE}
library(broom)
set.seed(17)
n <- 100
x <- rbinom(n, size = 1, prob = 0.5)
y <- ifelse(x==1, rbinom(n, size = 1, prob = 0.7), rbinom(n, size = 1, prob = 0.2))
o <- rnorm(n)
df <- data.frame(instrument = x, treatment = y, outcome = o)

cat("Part 1: cross tabulation of receipt of instrument and treatment\n")
table(instrument = x, treatment = y)
prop.table(table(instrument = x, treatment = y), 1)
lmfit <- lm(treatment ~ instrument, data = df)
tidy(lmfit)

cat("Part 2: linear regression of outcome on instrument\n")
lmfit2 <- lm(outcome ~ instrument, data = df)
tidy(lmfit2)

cat("Part 3: linear regression of outcome on predicted treatment\n")
df$predicted_trt <- predict(lmfit)
lmfit3 <- lm(outcome ~ predicted_trt, data = df)
tidy(lmfit3)

cat("Part 4\n")
a <- tidy(lmfit2)$estimate[2]
b <- tidy(lmfit)$estimate[2]
cat(a, "/", b, "=", a/b, "\n")
```

Example 2

```{r echo=FALSE}
set.seed(42)
n <- 100
x <- rbinom(n, size = 1, prob = 0.5)
y <- ifelse(x==1, rbinom(n, size = 1, prob = 0.7), rbinom(n, size = 1, prob = 0.2))
o <- rnorm(n)
df <- data.frame(instrument = x, treatment = y, outcome = o)

cat("Part 1: cross tabulation of receipt of instrument and treatment\n")
table(instrument = x, treatment = y)
prop.table(table(instrument = x, treatment = y), 1)
lmfit <- lm(treatment ~ instrument, data = df)
tidy(lmfit)

cat("Part 2: linear regression of outcome on instrument\n")
lmfit2 <- lm(outcome ~ instrument, data = df)
tidy(lmfit2)

cat("Part 3: linear regression of outcome on predicted treatment\n")
df$predicted_trt <- predict(lmfit)
lmfit3 <- lm(outcome ~ predicted_trt, data = df)
tidy(lmfit3)

cat("Part 4\n")
a <- tidy(lmfit2)$estimate[2]
b <- tidy(lmfit)$estimate[2]
cat(a, "/", b, "=", a/b, "\n")
```

Example 3:

```{r echo=FALSE}
set.seed(76)
n <- 100
x <- rbinom(n, size = 1, prob = 0.5)
y <- ifelse(x==1, rbinom(n, size = 1, prob = 0.7), rbinom(n, size = 1, prob = 0.2))
o <- rnorm(n)
df <- data.frame(instrument = x, treatment = y, outcome = o)

cat("Part 1: cross tabulation of receipt of instrument and treatment\n")
table(instrument = x, treatment = y)
prop.table(table(instrument = x, treatment = y), 1)
lmfit <- lm(treatment ~ instrument, data = df)
tidy(lmfit)

cat("Part 2: linear regression of outcome on instrument\n")
lmfit2 <- lm(outcome ~ instrument, data = df)
tidy(lmfit2)

cat("Part 3: linear regression of outcome on predicted treatment\n")
df$predicted_trt <- predict(lmfit)
lmfit3 <- lm(outcome ~ predicted_trt, data = df)
tidy(lmfit3)

cat("Part 4\n")
a <- tidy(lmfit2)$estimate[2]
b <- tidy(lmfit)$estimate[2]
cat(a, "/", b, "=", a/b, "\n")
```
