---
title: "Introduction to survival data"
output:
  ioslides_presentation:
    widescreen: true
    css: ../styles/styles.css
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.align="center")
```

## Studying long-term disease

We enroll people in a study with and without an exposure (history of obesity).

We want to see whether this exposure affects development of an outcome of interest (cancer).

What complications may arise when collecting data?

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = 0, y0 = 5:3, x1 = 1, lwd = 4)
points(x = c(0,0,0,1), y = 5:2, pch = 16, cex = 2)
abline(v = c(1,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1), y0 = 5:2, x1 = 2, lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
abline(v = c(2,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = 3, lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
abline(v = c(3,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,rep(4,4)), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = rep(4,2), y = c(4,3), pch = "X", cex = 2)
abline(v = c(4,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,5,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(5,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,6,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(6,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,7,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = 3, y = 5, pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(7,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,8,5), lwd = 4)
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = c(3,8), y = c(5,2), pch = 1, cex = 2)
points(x = c(rep(4,2),5), y = c(4,3,1), pch = "X", cex = 2)
abline(v = c(8,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

----

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,8,5), lwd = 4)
segments(x0 = 8, y0 = 2, x1 = 10, lwd = 4, lty = "dashed", col = "gray50")
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = c(3,8), y = c(5,2), pch = 1, cex = 2)
points(x = c(rep(4,2),5,10), y = c(4,3,1,2), pch = "X", cex = 2)
abline(v = c(8,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
```

## Time-to-event data

Studies in which patient outcomes are followed over time tend to use analysis methods in a statistical area known as **survival analysis**.

Key feature: outcome is the time to an event. This time is called a **survival time**, and the event is typically called a **failure**.

- Data collected from such studies almost always exhibit **censoring**.
- Reflect: how should time be defined? Age, time since study began?

## Examples of time-to-event data

- Time to hospital readmission
- Time to reincarceration
- Time to understanding of a concept

## Survival analysis vs. other approaches

Why not treat failure as a binary event and use logistic regression?

- Ignores time information

Wny not treat the time-to-event as a continous variable and use linear regression?

- Ignores censoring information

## Goals of survival analysis

- Estimate and compare survival over time (survival functions) for groups of interest
- Estimate the effect of covariates on survival

----

<h2 class="divider">Censoring</h2>

## Types of censoring

- Right-censoring
- Left-censoring
- Interval-censoring

## Right-censoring

True survival time is greater than or equal to the observed survival time.

- Data collection stops at a prespecified time.
- Subject withdraws from study or is lost to follow-up before the failure.

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,6), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = c(0,0,0,1,2), y0 = 5:1, x1 = c(3,4,4,8,5), lwd = 4)
segments(x0 = 8, y0 = 2, x1 = 10, lwd = 4, lty = "dashed", col = "gray50")
points(x = c(0,0,0,1,2), y = 5:1, pch = 16, cex = 2)
points(x = c(3,8), y = c(5,2), pch = 1, cex = 2)
points(x = c(rep(4,2),5,10), y = c(4,3,1,2), pch = "X", cex = 2)
abline(v = c(8,8), col = c("red", "blue"), lty = "dashed")
text(x = 7, y = 0.25, labels = "End of study", col = "blue")
text(x = 2.25, y = 5.25, labels = "Dropout")
text(x = c(1.75,6.75), y = c(4.75,1.75), labels = "Right-censored", col = "red")
```

## Left-censoring

True survival time is less than or equal to the observed survival time.

- Disease detection occurs after physiological onset. e.g. Flu infection has occurred before you start to get a fever.

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,2), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = 0, x1 = 7, y0 = 1, lwd = 4)
points(x = c(5,7), y = c(1,1), pch = "X", cex = 2, col = c("red", "black"))
text(x = c(5,7), y = c(0.75,1.25), labels = c("Infection onset", "Infection detection"), col = c("red", "black"))
```

## Interval censoring

True survival time is between two known times

- Disease is detected between regular follow up visits

```{r}
par(mar = c(4.5,4,0.2,1))
plot(1, type = "n", xlim = c(0,10), ylim = c(0,2), yaxt = "n", xlab = "Year", ylab = "Individuals")
segments(x0 = 0, x1 = 7, y0 = 1, lwd = 4)
points(x = c(3,5,7), y = rep(1,3), pch = "X", cex = 2, col = c("black", "red", "black"))
text(x = c(3,5,7), y = c(1.25,0.75,1.25), labels = c("No infection detection", "Infection onset", "Infection detection"), col = c("black", "red", "black"))
```

----

<h2 class="divider">Estimating survival quantities</h2>

## Notation and terminology

- $T_i$: observed survival time for subject $i$
- $\Delta_i$: censoring indicator for subject $i$ (1 = censored, 0 = observed)
- Data in a survival analysis setting include $T_i, \Delta_i$ and covariates.

## Functions of interest

Survivor function: probability of survival past a time $t$

$$ S(t) = P(T > t) = 1 - F(t) $$

Hazard function: instantaneous rate of failure at time $t$ given survival up to time $t$

$$ h(t) = \lim_{\Delta t \to 0} \frac{P(t \leq T < t+\Delta t \mid T \geq t)}{\Delta t} $$

## Visualizing functions of interest

```{r fig.width=10}
par(mfrow = c(1,2))
t <- seq(0, 20, 0.01)
s1 <- pexp(t, lower.tail = FALSE, rate = 0.1)
s2 <- pexp(t, lower.tail = FALSE, rate = 0.2)
plot(t, s1, type = "l", ylim = range(s1, s2), xlab = "Time", ylab = "Survival probability")
lines(t, s2, col = "red")
text(x = c(5,15), y = c(0.05,0.45), labels = c("Worse survival", "Better survival"), col = c("red", "black"))
## Hazard
plot(t, rep(0.1, length(t)), type = "l", xlab = "Time", ylab = "Hazard", ylim = c(0,0.3))
lines(t, rep(0.2, length(t)), col = "red")
text(x = c(10,10), y = c(0.07,0.23), labels = c("Better survival", "Worse survival"), col = c("black", "red"))
```

## Visualizing hazard functions

```{r fig.width=10}
par(mfrow = c(2,2), mar = c(4.1,4.1,3.1,1))
h1 <- rep(3, length(t))
h2 <- exp(t-10)
h3 <- exp(-t-2)
h4 <- sin(t*1.5*pi/20)
plot(t, h1, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Healthy individuals")
plot(t, h2, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Cancer patients")
plot(t, h3, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Newborn infants")
plot(t, h4, type = "l", xlab = "Time", ylab = "Hazard", xaxt = "n", yaxt = "n", main = "Acute onset, long-term disease")
```

## Goal: estimate and compare functions among groups

```{r fig.width=10}
library(survival)
data(lung)
lung$SurvObj <- with(lung, Surv(time, status == 2))
km.by.sex <- survfit(SurvObj ~ sex, data = lung, conf.type = "log-log")
t <- seq(0,1000)
s_true1 <- pexp(t, lower.tail = FALSE, rate = 0.002)
s_true2 <- pexp(t, lower.tail = FALSE, rate = 0.004)
par(mfrow = c(1,2))
plot(t, s_true1, type = "l", xlab = "Time from study enrollment (days)", ylab = "Survival probability", main = "True survival functions", ylim = c(0,1))
lines(t, s_true2, col = "red")
legend("topright", legend = c("group 1", "group 2"), col = c("black", "red"), lty = "solid", bty = "n")
plot(km.by.sex, mark.time = FALSE, col = c("red", "black"), xlab = "Time from study enrollment (days)", main = "Estimated survival functions")
```

## Estimating the survival function

- The **Kaplan-Meier method** is the most common method used to estimate survival functions.
- Accounts for right-censoring in survival data.
- In-class example
